---
title: "HCABrowser"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Connecting to the Human Cell Atlas  
The `HCABrowser` package relies on having network connectivety. Also, the HCA's Data Coordination Platform (DCP) must also be operational.  

The `HCABrowser` object serves as the representation of the Human Cell Atlas. Upon creation, it will automatically peform a cursorary query and display a small table showing the first few bundles of the entire HCA. This intial table contains some columns that we have determined are most useful to users. The output also displays the url of the instance of the HCA DCP being used, the current query, whether bundles or files are being displayed, and the number of bundles in the results.  

By default, ten bundles per page will be displayed in the result and the default url to the HCA DCP will be used. Currently, the maximum number of bundles that can be shown per query is ten. These two values can be changed in the constructor.  

If the HCA cannot be reached, an error will be thrown displaying the status of the request.  


```{r}
hca <- HCABrowser(url = "https://dss.data.humancellatlas.org/v1", 
                  per_page = 10)
```

