---
title: "GenomeInfoDb"
author: "yincy"
date: "11/30/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
```

```{r}
library(GenomeInfoDb)
library(GenomicRanges)
library(magrittr)
ls("package:GenomeInfoDb")
```


## Introduction  
The `GenomeInfoDb` provides an interface to access `seqlevelsStyles` (such as UCSC, NCBI, Ensembl) and their supported mappings for organisms. For instance, for Homo sapiens, `seqlevelsStyle` "UCSC" maps to "chr1", "chr2", ..., "chrX","chrY".  


##Functionality for all existing organisms   
### genomeStyles  
The `genomeStyles` lists out for each organism, the `seqlevelsStyles` and their mappings.  

```{r}
seqmap <- genomeStyles()
seqmap %>% names()
seqmap[["Mus_musculus"]]
```


Organism's supported by `GenomeInfoDb` can be found by  
```{r}
names(genomeStyles())
```

If one knows the organism one is interested in, then we can directly access the information for the given organism along.  

```{r}
genomeStyles(species = "Homo_sapiens")
```

We can also check if a given style is supported by `GenomeInfoDb` for a given species.  
```{r}
"UCSC" %in% names(genomeStyles(species = "Homo_sapiens")) 
```


### extractSeqlevels   
We can also extract the desired `seqlevelsStyle` from a given organism using the `extractSeqlevels`  
```{r}
extractSeqlevels(species = "Homo_sapiens", style = "NCBI")
```

### extractSeqlevelsByGroup   
We can also extract the desired `seqlevelsStyle` from a given organism based on a group ('auto' denotes autosomes, 'circular' denotes circular chromosomes and 'sex' denotes sex chromosomes, the default is all chromosomes are returned).  

```{r}
extractSeqlevelsByGroup(species = "Homo_sapiens", group = "all", style = "NCBI")
```


### seqlevelsStyle  
We can find the seqname Style for a given character vector by using `seqlevelsStyle`  
```{r}
seqlevelsStyle(x = paste("chr", 1:30, sep = ""))
```

```{r}
seqlevelsStyle(x = c("2L", "2R", "X", "Xhet"))
```


### seqlevelsInGroup  
We can also subset a given character vector containing seqnames using the `seqlevelsInGroup`.  

We currently support 3 groups: 'auto' for autosomes, 'sex' for all or some sex chromosomes and 'circular' for 'circular' chromosomes.  
```{r}
newchr <- paste0("chr",c(1:22,"X","Y","M","1_gl000192_random","4_ctg9_hap1"))

seqlevelsInGroup(newchr, group = "sex")
```

```{r}
seqlevelsInGroup(seqnames = newchr,
                 group = "auto")
```



```{r}
seqlevelsInGroup(seqnames = newchr, group = "circular")
```

```{r}
seqlevelsInGroup(seqnames = newchr, 
                 group = "sex", 
                 species = "Homo_sapiens", 
                 style = "UCSC")
```



### orderSeqlevels  
The `orderSeqlevels` can return the order of a given character vector which contains seqnames.  
```{r}
seqnames <- c("chr1", "chr9", "chr2", "chr3", "chr10")
orderSeqlevels(seqnames)
```

```{r}
seqnames[orderSeqlevels(seqnames)]
```

### rankSeqlevels  
The `rankSeqlevels` can return the rank of a given character vector which contains seqnames.  
```{r}
seqnames <- c("chr1","chr9", "chr2", "chr3", "chr10")
rankSeqlevels(seqnames = seqnames)
seqnames[rankSeqlevels(seqnames)]
```


### mapSeqlevels  
Returns a matrix with 1 column per supplied sequence name and 1 row per sequence renaming map compatible with the specified style.  

If the `best.only` is TRUE (defualt), only the 'best' renaming maps are returned  
```{r seqlevel transformation}
mapSeqlevels(seqnames = c("chrII", "chrIII", "chrM"), style = "NCBI")
```


```{r}
gr <- GRanges(seqnames = paste('ch', 1:35, sep = ""), 
              ranges = IRanges(start = 1:35, width = 5))
gr %>% seqlevels()
seqnames(gr) <- mapSeqlevels(seqnames = as.character(seqnames(gr)), style = "NCBI") %>% unname() %>% as.factor() %>% Rle
```


### renameSeqlevels  
```{r}
newnames <- paste("chr", 1:35, sep = "")
names(newnames) <- paste("ch", 1:35)
head(newnames)
```

```{r}
gr <- renameSeqlevels(x = gr, value = newnames)
```


### dropSeqlevels  
Here the second argument is the seqlevels that you want to drop. Because these seqlevels are in use (i.e. have ranges on them), the ranges on these sequences need to be removed before the seqlevels can be dropped.   

We call this pruning. The `pruning.mode` argument controls how to prune `gr`. Unlike for list-like objects (e.g. GRangesList) for which pruning can be done in various ways, pruning a GRanges object is straightforward and achieved by specifying `pruning.mode="coarse"`.  

```{r}
dropSeqlevels(x = gr, 
              value = paste("chr", 23:35, sep = ""), 
              pruning.mode = "coarse")
```

### keepSeqlevels  
```{r}
keepSeqlevels(x = gr, value = paste0("chr", 1:22), pruning.mode = "coarse")
```


### keepStandardChromosomes  
This function internally uses the pre-defined tables inside GenomeInfoDb to find the correct seqlevels according to the sequence style of the object.  

```{r}
keepStandardChromosomes(x = gr, pruning.mode = "coarse")
```


```{r}
plantgr <- GRanges(seqnames = c(1:5, "MT", "Pltd"), 
                   ranges = IRanges(start = 1:7, width = 5))

keepStandardChromosomes(x = plantgr, 
                        species = "Arabidopsis thaliana", 
                        pruning.mode = "coarse")
```


## Classes inside GenomeInfoDb package  
### Genome-description class  
```{r}
library(BSgenome.Hsapiens.UCSC.hg38)
class(Hsapiens)
```

```{r}
is(object = Hsapiens, class2 = "GenomeDescription")
```


```{r}
provider(Hsapiens)
```


```{r}
seqinfo(Hsapiens)
```


```{r}
gendesc <- as(object = Hsapiens, Class = "GenomeDescription")
```

```{r}
class(gendesc)
```

```{r}
provider(gendesc)
```


```{r}
seqinfo(gendesc)
```

```{r}
bsgenomeName(gendesc)
```


### Seqinfo class  
Note that all the arguments (except 'genome') must have the same length. 'genome' can be of length 1, whatever the lengths of the other arguments are.  
```{r}
x <- Seqinfo(seqnames = c("chr1", "chr2", "chr3", "chrM"),
             seqlengths = c(100, 200, NA, 15),
             isCircular = c(NA, FALSE, FALSE, TRUE),
             genome = "toy")
```


```{r}
length(x)
```

```{r}
seqnames(x)
```

```{r}
names(x)
```

```{r}
seqlevels(x)
```


```{r}
isCircular(x)
```


```{r}
genome(x)
```

```{r subset by name}
x[c("chrY", "chr3", "chr1")]
```


```{r rename seq names}
xx <- x
seqlevels(xx) <- sub("chr", "ch", seqlevels(x))
```

```{r}
seqlevels(xx) <- rev(seqlevels(xx))
xx
```

```{r drop/add/reorder}
seqlevels(xx) <- c("ch1", "ch2", "chY")
xx
```


```{r rename/reorder/drop/add}
seqlevels(xx) <- c(chY = "Y", ch1 = "1", "22")
xx
```


```{r}
y <- Seqinfo(seqnames = c("chr3", "chr4", "chrM"),
             seqlengths = c(300, NA, 15))
```


```{r}
merge(x, y)
```


```{r}
merge(y, x)
```

Note that, strictly speaking, merging 2 Seqinfo objects is not a commutative operation, i.e., in general 'z1 <- merge(x, y)' is not identical to 'z2 <- merge(y, x)'. However 'z1' and 'z2' are guaranteed to contain the same information (i.e. the same rows, but typically not in the same order) 














