---
title: "IRanges"
author: "yincy"
date: "11/5/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load package, message=FALSE}
library(IRanges)
library(magrittr)
```

Two central classes are available in Bioconductor for representing ranges: the **IRanges** class defined in the `IRanges` package for representing ranges defined on a single space, and the `GRanges` class defined in the `GenomicRanges` package for representing ranges defined on multiple spaces.  


`IRanges` are collections of integer intervals. `GRanges` are like `IRanges`, but with an associated chromosome and strand, taking care of some book keeping.  


## Basic IRanges  
```{r}
ir1 <- IRanges(start = 1:10, width = 10:1)
```

```{r}
ir2 <- IRanges(start = 1:10, width = 11)
ir3 <- IRanges(end = 11, width = 10:1)
```

```{r}
identical(ir1, ir2)
```

An `IRanges` consist of separate intervals; each interval is called a range.  

**Assessor methods: `start()`, `end()`, `width()` and also replacement methods**.  

```{r}
start(ir1)
```

```{r}
width(ir1)
```

```{r}
width(ir1) <- 1
ir1
```

```{r give they are names}
names(ir1) <- paste("A", 1:length(ir1), sep = "") # row names 
ir1
```

```{r IRange object only have one dimension}
dim(ir1); length(ir1)
```

```{r subsetting like vector}
ir1[1]
```

```{r}
ir1["A1"]
```

```{r concatenate two IRanges with c() function}
c(ir1, ir2)
```

```{r plotRanges}
plotRanges <- function(x, 
                       xlim = x, 
                       main = deparse(substitute(x)), 
                       col = "black", 
                       sep = 0.3, 
                       ...){
    height <- 1
    if(is(xlim, "IntegerRanges")){
        xlim <- c(min(start(xlim)), max(end(xlim)))
    }
    bins <- disjointBins(IRanges(start(x), end(x) + 1))
    plot.new()
    plot.window(xlim = xlim, ylim = c(0, max(bins) * (height + sep)))
    ybottom <- bins * (sep + height) - height
    rect(xleft = start(x) - 0.5, 
         ybottom = ybottom, 
         xright = end(x) + 0.5, 
         ytop = ybottom + height, 
         col = col, 
         ...)
    title(main)
    axis(1)
    axis(2)
                       }

plotRanges(ir3)
```


## Normality  
Sometimes, it is necessary to formally represent a subsequence, where no elements are repeated and order is preserved. Also, it is occasionally useful to think of an `IRanges` object as a set of integers, where no elements are repeated and order does not matter.  

The *NormalIRanges* class formally represents a set of integers. By definition an *IRanges* object is said to be normal when its ranges are: (a) not empty. (b) not overlapping. (c) ordered from left to right. (d) not even adjacent (e.g., there must be a non empty gap between 2 consecutive ranges.)

There are three main advantages of using a normal `IRanges` object:  

1. it guarantees a subsequence encoding or set of integers.  
2. it is compact in terms of the number of ranges.  
3. it uniquely identifies its informations, which simplifies comparisions.   

The `reduce` function reduces any `IRanges` object to a `NormalIRanges` by merging redundant ranges.  

```{r}
ir <- IRanges(start = c(1, 3, 5, 7), 
              end = c(2, 3, 8, 10))

par(mfrow = c(2, 1))
plotRanges(ir)
plotRanges(reduce(ir))
```


## Lists of IRanges objects  
It is common to manipulate collections of `IRanges` objects during an analysis. Thus, the `IRanges` package defines some specific classes for working with multiple `IRanges` objects.  

The `IRangesList` class asserts that each element is an `IRanges` object and provides convenience methods, such as `start`, `end` and `width` accessors that return `IntegerList` objects, aligning with the `IRangesList` object.  

To explicitly construct an `IRangesList`, use the `IRangesList` function.  

```{r}
rl <- IRangesList(ir, rev(ir))
start(rl)
```

## Vector Extraction  
> IRange object can be used for vector extraction.  

As the elements of an `IRanges` object encode consecutive subsequences, they may be used directly in sequence extraction. Note that when a normal `IRanges` is given as the index, the result is a subsequence, as no elements are repeated or reordered. If the sequence is a Vector subclass (i.e. not an ordinary vector), the canonical `[` function accepts an `IRanges` object.  

```{r}
set.seed(0)
lambda <- c(rep(0.001, 4500), 
            seq(0.001, 10, length = 500), 
            seq(10, 0.001, length = 500))

xVector <- rpois(1e7, lambda = lambda)
yVector <- rpois(1e7, lambda = lambda[c(251:length(lambda), 1:250)])
xRle <- Rle(xVector)
yRle <- Rle(yVector)
irextract <- IRanges(start = c(4501, 4901), width = 100)
xRle[irextract]
```

## Finding Overlapping Ranges  
The function `findOverlaps` detects overlaps between two `IRanges` objects.  

```{r}
ol <- findOverlaps(query = ir, reduce(ir))
as.matrix(ol)
ir[queryHits(ol)]
```

## Counting Overlapping Ranges  
The function `coverage` counts the number of ranges over each position.  

```{r}
cov <- coverage(ir) %>% as.vector()
mat <- cbind(seq_along(cov) - 0.5, cov)
d <- diff(cov) != 0 
mat <- rbind(cbind(mat[d, 1] + 1, mat[d, 2]), mat)
mat <- mat[order(mat[, 1]), ]

plotRanges(ir)
lines(mat, col = "red", lwd = 2)
```





