---
title: "monocle3-Differential-expression-analysis"
author: "yincy"
date: "4/8/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(monocle3)
library(tidyverse)
```

Differential gene expression analysis is a common task in RNA-Seq experiments. Monocle can help you find genes that are differentially expressed between groups of cells and assesses the statistical signficance of those changes. Monocle 3 includes a powerful system for finding genes that vary across cells of different types, were collected at different developmental time points, or that have been perturbed in different ways.  

There are two approaches for differential analysis in Monocle  

- **Regression analysis**: using `fit_models()`, you can evaluate whether each gene depends on variables such as time, treatments, etc.  

- **Graph-autocorrelation analysis**: using `graph_test()`, you can find genes that vary over a trajectory or between clusters.  

Monocle also comes with specialized functions for finding co-regulated modules of differentially expressed genes. Monocle also allows you to interactively interrogate specific clusters or regions of a trajectory (e.g. branch points) for genes that vary within them.  

### Regression analysis  
In this section, we'll explore how to use Monocle to find genes that are differentially expressed according to several different criteria. Performing differential expression analysis on all genes in a cell_data_set object can take anywhere from minutes to hours, depending on how complex the analysis is. To keep the vignette simple and fast, we'll be working with small sets of genes. Rest assured, however, that Monocle can analyze several thousands of genes even in large experiments, making it useful for discovering dynamically regulated genes during the biological process you're studying.  

Let's begin with a small set of genes that we know are important in ciliated neurons to demonstrate Monocle's capabilities.  



```{r}
load(file = "monocle3-Constructing-single-cell-trajectorie.RData")

ciliated_genes <- c(
    "che-1", 
    "hlh-17", 
    "nhr-6", 
    "dmd-6", 
    "ceh-36", 
    "ham-1"
)

cds_subset <- cds[rowData(cds)$gene_short_name %in% ciliated_genes, ]
```

The differential analysis tools in Monocle are extremely flexible. Monocle works by fitting a regression model to each gene. You can specify this model to account for various factors in your experiment (time, treatment, and so on). For example, In the embryo data, the cells were collected at different time points. We can test whether any of the genes above change over time in their expression by first fitting a generalized linear model to each one.  

log(y~i~) = β~0~ + β~t~x~t~

where y~i~ is a random variable corresponding to the expression values of gene i, x~t~ is the time each cell was collected (in minutes), and the β~t~ capture the effect of time on expression, and β~0~ is an intercept term. We can identify genes that vary over time by fitting this model to each one, and then testing whether it's β~t~ is significantly different from zero. To do so, we firt call the `fit_models()` function.  
```{r}
gene_fits <- fit_models(cds = cds_subset, 
                        model_formula_str = "~embryo.time")
```

`gene_fits` is a tibble that contains a row for each gene. The `model` column contains generalized linear model objects, each of which aims to explain the expression of a gene across the cells using the equation above.   

The parameter `model_formula_str` should be a string specifying the model formula. The model formulae you use in your tests can include any term that exists as a column in the colData table, including those columns that are added by Monocle in other analysis steps. For example, if you use cluster_cells, you can test for genes that differ between clusters and partitions by using` ~cluster` or `~partition` (respectively) as your model formula. You can also include multiple variables, for example `~embryo.time + batch`, which can be very helpful for subtracting unwanted effects.  

Now let's see which of these genes have time-dependent expression. First, we extract a table of coefficients from each model using the `coefficient_table()` function.  
```{r}
fit_coefs <- coefficient_table(gene_fits)
```

Note that the table includes one row for each term of each gene's model. We generally don't care about the intercept term β~0~, so we can easily just extract the time terms.  
```{r}
emb_time_terms <- fit_coefs %>% 
    dplyr::filter(term == "embryo.time")

emb_time_terms %>% 
    dplyr::filter(q_value < 0.05) %>% 
    dplyr::select(gene_short_name, term, q_value, estimate)
```

```{r, message=FALSE, warning=FALSE}
plot_genes_violin(
    cds_subset = cds_subset, 
    group_cells_by = "embryo.time.bin",
    ncol = 2
) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


### Controlling for batch effects and other factors  
```{r}
gene_fits <- fit_models(cds = cds_subset, 
                        model_formula_str = "~embryo.time + batch")

fit_coefs <- coefficient_table(gene_fits)
fit_coefs %>% 
    dplyr::filter(term != "(Intercept)") %>% 
    dplyr::select(gene_short_name, term, q_value, estimate)
```

### Evaluating models of gene expression  
How good are these models at "explaining" gene expression? We can evaluate the fits of each model using the `evaluate_fits()` function.  
```{r}
evaluate_fits(gene_fits)
```

Should we include the batch term in our model of gene expression or not? Monocle provides a function compare_models() that can help you decide. Compare models takes two models and returns the result of a likelihood ratio test between them. Any time you add terms to a model, it will improve the fit. But we should always to use the simplest model we can to explain our data. The likelihood ratio test helps us decide whether the improvement in fit is large enough to justify the complexity our extra terms introduce.   

```{r}
time_batch_models <- fit_models(
    cds = cds_subset, 
    model_formula_str = "~embryo.time + batch", 
    expression_family = "negbinomial"
)

time_models <- fit_models(
    cds = cds_subset, 
    model_formula_str = "~embryo.time", 
    expression_family = "negbinomial"
)

compare_models(model_tbl_full = time_batch_models, model_tbl_reduced = time_models) %>% 
    dplyr::select(gene_short_name, q_value)
```

The first of the two models is called the full model. This model is essentially a way of predicting the expression value of each gene in a given cell knowing both what time it was collected and which batch of cells it came from. The second model, called the reduced model, does the same thing, but it only knows about the time each cell was collected. Because the full model has more information about each cell, it will do a better job of predicting the expression of the gene in each cell. The question Monocle must answer for each gene is how much better the full model's prediction is than the reduced model's. The greater the improvement that comes from knowing the batch of each cell, the more significant the result of the likelihood ratio test.  

As we can see, **all of the gene's likelihood ratio tests are significant**, indicating that there are substantial batch effects in the data. We are therefore justified in adding the `batch` term to our model.  


### Choosing a distribution for modeling gene expression  
Monocle uses generalized linear models to capture how a gene's expression depends on each variable in the experiment. These models require you to specify a distribution that describes gene expression values. Most studies that use this approach to analyze their gene expression data use the negative binomial distribution, which is often appropriate for sequencing read or UMI count data. The negative binomial is at the core of many packages for RNA-seq analysis, such as DESeq2.  

Monocle's `fit_models()` supports the negative binomial distribution and several others listed in the table below. The default is the "quasipoisson", which is very similar to the negative binomial. Quasipoisson is a a bit less accurate than the negative binomial but much faster to fit, making it well suited to datasets with thousands of cells.  

There are several allowed values for expression_family  

**expression_family** | **Distribution**   | **Accuracy**  | **Speed** | **Notes**  
----------------------|--------------------|---------------|-----------|-----------------
`quasipoisson`        | Quasi-poisson      | ++            | ++        | Default for `fit_models()`. recommended for most users.  
`negbinomial`         | negative binomial  | +++           | +         | Recommended for users with small datasets (< 1000 cells)
`poisson`             | Poisson            | -             | +++       | Not recommended. For debugging and testing only.  
`binomial`            | Binomial           | ++            | ++        | recommended for single-cell ATAC-seq  


**Likelihood based analysis and quasipoisson**  
The quasi-position distribution doesn't have a real likelihood function, so some of Monocle's methods won't work with it. Serveral of the columns in results tables from `evaluate_fits()` and `compare_models()` will be `NA`.  


### Graph-autocorrelation analysis for comparing clusters  
In the L2 worm data, we identified a number of clusters that were very distinct as neurons.  
```{r}
expression_matrix <- readRDS(file = "f:/git/Data/monocle3/cao_l2_expression.rds")
cell_metadata <- readRDS(file = "f:/git/Data/monocle3/cao_l2_colData.rds")
gene_metadata <- readRDS(file = "f:/git/Data/monocle3/cao_l2_rowData.rds")

cds <- new_cell_data_set(
    expression_data = expression_matrix, 
    cell_metadata = cell_metadata, 
    gene_metadata = gene_metadata
)

neurons_cds <- cds[, grepl("neurons", colData(cds)$cao_cell_type, ignore.case = T)]

neurons_cds <- preprocess_cds(cds = neurons_cds, 
                      num_dim = 50, 
                      method = "PCA", 
                      norm_method = "log")

neurons_cds <- align_cds(cds = neurons_cds, 
                 preprocess_method = "PCA", 
                 alignment_group = "plate")

neurons_cds <- reduce_dimension(cds = neurons_cds, 
                        max_components = 2, 
                        reduction_method = "UMAP", 
                        preprocess_method = "PCA")

neurons_cds <- cluster_cells(cds = neurons_cds, 
                             reduction_method = "UMAP", 
                             k = 20)

plot_cells(cds = neurons_cds, 
           reduction_method = "UMAP", 
           color_cells_by = "partition")
```

There are many subtypes of neurons, so perhaps the different neuron clusters correspond to different subtypes. To investigate which genes are expressed differentially across the clusters, we could use the regression analysis tools discussed above. However, Monocle provides an alternative way of finding genes that vary between groups of cells in UMAP or t-SNE space. The function graph_test() uses a statistic from spatial autocorrelation analysis called Moran's I, which Cao & Spielmann et al showed to be effective in finding genes that vary in single-cell RNA-seq datasets.  

run `graph_test()`  
```{r}
pr_graph_test_res <- graph_test(cds = neurons_cds, 
                                neighbor_graph = "knn", 
                                cores = 8)

pr_deg_ids <- row.names(subset(pr_graph_test_res, q_value < 0.05))
```

The data frame pr_graph_test_res has the Moran's I test results for each gene in the cell_data_set. If you'd like to rank the genes by effect size, sort this table by the morans_Icolumn, which ranges from -1 to +1. A value of 0 indicates no effect, while +1 indicates perfect positive autocorrelation and suggests that a nearby cells have very similar values of a gene's expression. Significant values much less than zero are generally rare.  

Positive values indicate a gene is expressed in a focal region of the UMAP space (e.g. specific to one or more clusters). But how do we associate genes with clusters? The next section explains how to collect genes into modules that have similar patterns of expression and associate them with clusters.  


### Finding modules of co-regulated genes  





























