---
title: "GSEABase"
author: "yincy"
date: "4/12/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Abstract  
The `GSEABase` package implements data structures and methods to represent, manipulate,and analyze gene sets in the context of gene set enrichment analysis. This includes construction of gene sets from reference resources, ID mapping, coloring according to phenotype association, and storing in gene set collections.   


# GeneSet  
A `GeneSet` stores a set of related gene identifiers. Important components of the gene set are a vector of identifiers, general descriptive information about the set, and information about how the gene set was constructed. To construct a gene set, use `GeneSet`. For example, to create a gene set from the identifiers in a subset of the sample *ExpressionSet* in the Biobase package use.  


```{r, message=FALSE}
library(GSEABase)
data(sample.ExpressionSet, package = "Biobase")

sample.ExpressionSet
```

```{r}
egs <- GeneSet(sample.ExpressionSet[201:250, ], setName = "sample")
```

```{r}
details(egs)
```

```{r}
showMethods("GeneSet", inherited = F)
```

The identifier type of gene sets created from expression sets is AnnotationIdentifier. Additional predefined identifiers are available.  
```{r}
names(slot(getClass("GeneIdentifierType"), "subclasses"))
```



```{r}
egss <- GeneSet(setName = "sample", 
                setIdentifier = rownames(sample.ExpressionSet),  
                geneIds = "ensembl", 
                organism = "hsapiens", 
                pubMedIds = "123446", 
                urls = "www.yincy.cn")
```















