---
title: "GSEABase"
author: "yincy"
date: "3/25/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Abstract  
The GSEABase package implements data structures and methods to represent, manipulate, and analyze gene sets in the context of gene set enrichment analysis. This includes construction of gene sets from reference resources, ID mapping, coloring according to phenotype association, and storing in gene set collections.  

## GeneSet  
A GeneSet stores a set of related gene identifiers. Important components of the gene set are a vector of identifiers, general description about the set, and information about how the gene set was constructed. To construct a gene set, use `GeneSet`. For example, to create a gene set from the identifiers in a subset of the sample ExpressionSet in the Biobase package use  
```{r}
library(GSEABase)
data(sample.ExpressionSet, package = "Biobase")
egs <- GeneSet(type = sample.ExpressionSet[201:250, ], setName = "sample")
```

Each gene set may have a name. The gene set contains 50 identifiers ('genes') from the ExpressionSet. These are accessible using `geneIds`  
```{r}
geneIds(egs) %>% head()
```

```{r}
details(egs)
```

The set identifier, set version, and creation date provide mechanisms for carefully curating gene sets. Additional information is automatically copied from the expression set used to create `egs`.  

View additional methods for creating gene sets with  
```{r}
showMethods(f = "GeneSet", inherited = T)
```

The identifier type of gene sets created from expression sets is AnnotationIdentifier. Addi-
tional predefined identifiers are available.  
```{r}
names(slot(getClass("GeneIdentifierType"), "subclasses"))
```

It is possible to map between identifier types  
```{r}
mapIdentifiers(egs, EntrezIdentifier(), verbose = T)
```



























