---
title: "fgsea"
author: "yincy"
date: "4/18/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


fgsea is an R-package for fast preranked gene set enrichment analysis (GSEA). The performance is achieved by using an algorithm for cumulative GSEA-statistic calculation. This allows to reuse samples between different gene set sizes.  


## Loading necessary libraries  
```{r, message=F, warning=FALSE}
library(fgsea)
library(tidyverse)
data(examplePathways) # a list 
data(exampleRanks)
```


```{r}
examplePathways[[1]]
exampleRanks %>% head()
```

```{r}
sapply(examplePathways, length) %>% sort(decreasing = F) %>% quantile(probs = seq(0, 1, 0.1))
```


```{r}
fgseaRes <- fgsea(pathways = examplePathways, # list of gene set to check 
                  stats = exampleRanks, 
                  minSize = 15, 
                  maxSize = 500,
                  nperm = 10000)
```


```{r}
fgseaRes %>% 
    arrange(pval)
```


```{r}
sum(fgseaRes[, padj < 0.01])
```


```{r}
plotEnrichment(pathway = examplePathways[["5990980_Cell_Cycle"]], 
               stats = exampleRanks, 
               ticksSize = 0.2) +
    labs(title = "Programmed Cell Death") +
    geom_vline(xintercept = 2000)
```


```{r}
topPathwaysUp <- fgseaRes[ES > 0, ][head(order(pval), n=10), pathway]
topPathwaysDown <- fgseaRes[ES < 0, ][head(order(pval), n=10), pathway]
topPathways <- c(topPathwaysUp, rev(topPathwaysDown))
plotGseaTable(pathways = examplePathways[topPathways], 
              stats = exampleRanks, 
              fgseaRes = fgseaRes,
              gseaParam = 0.5)
```


From the plot above one can see that there are very similar pathways in the table (for example 5991502_Mitotic_Metaphase_and_Anaphase and 5991600_Mitotic_Anaphase). To select only independent pathways one can use collapsePathways function:  
```{r}
collaspsedPathways <- collapsePathways(fgseaRes = fgseaRes[order(pval), ][padj < 0.01, ], 
                                       pathways = examplePathways, 
                                       stats = exampleRanks, 
                                       nperm = 1000)

mainPathways <- fgseaRes[pathway %in% collaspsedPathways$mainPathways, ][order(-NES), pathway]
plotGseaTable(examplePathways[mainPathways], exampleRanks, 
              fgseaRes = fgseaRes, 
              gseaParam = 0.5)
```



## Using Reactome pathways  
For convenience there is reactomePathways function that obtains pathways from Reactome for given set of genes. Package reactome.db is required to be installed.  

```{r}
pathways <- reactomePathways(genes = exampleRanks %>% names())
pathways %>% length()

sapply(pathways, length) %>% sort() %>% quantile(probs = seq(0, 1, 0.1))

fgseaRes <- fgsea(pathways = pathways, 
                  stats = exampleRanks, 
                  nperm = 1000, 
                  maxSize = 500, 
                  minSize = 10)

fgseaRes %>% arrange(padj) %>% head()
```






