---
title: "Integration and Label Transfer"
author: "yincy"
date: "3/5/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Introduction: Seurat v3 Integration  
As described in [Stuart*, Butler*, et al. Cell 2019](https://www.cell.com/cell/fulltext/S0092-8674(19)30559-8), Seurat v3 introduces new methods for the integration of multiple single-cell datasets. These methods aim to identify shared cell states that are present across different datasets, even if they were collected from different individuals, experimental conditions, technologies, or even species.  

Our method aims to first identify ‘anchors’ between pairs of datasets. These represent pairwise correspondences between individual cells (one in each dataset), that we hypothesize originate from the same biological state. These ‘anchors’ are then used to harmonize the datasets, or transfer information from one dataset to another. Below, we demonstrate multiple applications of integrative analysis, and also introduce new functionality beyond what was described in the 2019 manuscript.  

## Standard Workflow  

- Describes the standard Seurat v3 integration workflow, and applies it to integrate multiple datasets collected of human pancreatic islets (across different technologies). We also demonstrate how Seurat v3 can be used as a classifier, transferring cluster labels onto a newly collected dataset.  

- We recommend this vignette for new users  


## SCTransform  

- Describes a modification of the v3 integration workflow, in order to apply to datasets that have been normalized with our new normalization method, SCTransform. We apply this to the same pancreatic islet datasets as described previously, and also integrate human PBMC datasets from [eight different technologies](https://www.biorxiv.org/content/10.1101/632216v1), produced as a systematic technology benchmark by the Human Cell Atlas.  

- We recommend this vignette for advanced users who are familiar with our SCTransform normalization method. You can read more about SCTransform in our recent preprint, and see how to apply it to a single dataset in a separate vignette.  


## Reference-based  

- Describes a modification of the v3 integration workflow, where a subset of the datasets (or a single dataset) are listed as a ‘reference’. This approach can result in dramatic speed improvements, particularly when there are a large number of datasets to integrate. We apply this to the eight PBMC datasets described above, and observe identical results, despite a substantial reduction in processing time.  

- We recommend this vignette for users who are integrating many datasets, and are looking for speed improvements.  


## Reciprocal PCA  

- Describes a modification of the v3 integration workflow, where reciprocal PCA is used in place of canonical correlation analysis for the dimension reduction used in anchor finding. This approach can improve speed and efficiency when working with large datasets.  

- We recommend this vignette for users looking for speed/memory improvements when working with a large number of datasets or cells, for example experimental designs with many experimental conditions, replicates, or patients. However, this workflow may struggle to align highly divergent samples (e.g. cross species, or cross-modality, integration). For a ‘turbo’ mode, consider combining with “reference-based” integration as demonstrated here.  



# Standard Workflow  
In this example workflow, we demonstrate two new methods we recently introduced in our paper, [Comprehensive Integration of Single Cell Data](https://www.cell.com/cell/fulltext/S0092-8674(19)30559-8)  

- Assembly of multiple distinct scRNA-seq datasets into an integrated reference  
- Transfer of cell type labels from a reference dataset onto a new query dataset  

For the purposes of this example, we’ve chosen human pancreatic islet cell datasets produced across four technologies, CelSeq (GSE81076) CelSeq2 (GSE85241), Fluidigm C1 (GSE86469), and SMART-Seq2 (E-MTAB-5061).  

## Dataset preprocessing  
Load in the dataset. The metadata contains the technology (tech column) and cell type annotations (celltype column) for each cell in the four datasets.  
```{r, message=FALSE}
library(Seurat)
library(SeuratData)
library(tidyverse)
library(cowplot)
library(patchwork)
```

To construct a reference, we will identify 'anchors' between the individual datasets. First, we split the comnined object into a list, with each dataset as an element.  
```{r}
data("panc8")
pancrease.list <- SplitObject(object = panc8, split.by = "tech")
pancrease.list <- pancrease.list[c("celseq", "celseq2", "fluidigmc1", "smartseq2")]
```

Prior to finding anchors, we perform standard preprocessing (log-normalization), and identify variable features individually for each. Note that Seurat v3 implements an improved methods for variable feature selection based on a variance stabilizing transformation (`"vst"`).  
```{r}
for(i in 1:length(pancrease.list)){
    pancrease.list[[i]] <- NormalizeData(object = pancrease.list[[i]], verbose = F)
    pancrease.list[[i]] <- FindVariableFeatures(object = pancrease.list[[i]], 
                                                selection.method = "vst", 
                                                nfeatures = 2000, 
                                                verbose = F)
}
```


## Integration of 3 pancreatic islet cell datasets  
Next, we identify anchors using the `FindIntergrationAnchors` function, which takes a list of Seurat objects as input. Here, we integrate three of the objects into a reference.  

- We use all default parameters here for identifying anchors, including the 'dimensionality' of the dataset (30, try varying this parameter over a broad range, for example between 10 and 50).  
```{r}
reference.list <- pancrease.list[c("celseq", "celseq2", "smartseq2")]
pancrease.anchors <- FindIntegrationAnchors(object.list = reference.list, 
                                            dims = 1:30)
```

then pass these anchors to the `IntegrateData` function, which returns a Seurat object.  

- The returned object will contain a new `Assay`, which holds an integrated (or 'batch-corrected') expression matrix for all cells, enabling them to be jointly analyzed.  

```{r}
pancrease.integrated <- IntegrateData(anchorset = pancrease.anchors)
```

After running `IntegrateData`, the `Seurat` object will contain a new `Assay` with the integrated expression matrix. Note that the original (uncorrected values) are still stored in the object in the 'RNA' assay, so you can switch back and forth.  

We can then use this new integrated matrix for downstream analysis and visulization. Here we scale the integrated data, run PCA, and visualize the results with UMAP. The integrated datasets cluster by cell type, instead of by technology.  

```{r}
DefaultAssay(pancrease.integrated) <- "integrated"

# run the standard workflow for visalization and clustering  
pancrease.integrated <- ScaleData(object = pancrease.integrated, verbose = F)
pancrease.integrated <- RunPCA(object = pancrease.integrated, npcs = 30, verbose = F)
pancrease.integrated <- RunUMAP(object = pancrease.integrated, reduction = "pca", dims = 1:30)
```

```{r}
p1 <- DimPlot(object = pancrease.integrated, reduction = "umap", group.by = "tech")
p2 <- DimPlot(object = pancrease.integrated, reduction = "umap", group.by = "celltype", label = T, repel = T) + NoLegend()

plot_grid(p1, p2)
```

## Cell type classification using an integrated reference  
Seurat v3 also support the projection of reference data (or meta data) onto a query object. While many of the methods are conserved (both procedures begin by identify anchors), there are two important distintions between data transfer and integration:   

1. In data transfer, Seurat does not correct or modify the query expression data.  
2. In data transfer, Seurat has an option (set by default) to project the PCA structure of a reference onto the query, instead of learning a joint structure with CCA. We generally suggest using this option when projecting data between scRNA-seq datasets.  

After finding anchors, we use the `TransferData` function to classify the query cells based on reference data (a vector of reference cell type label). `TransferData` returns a matrix with predicted IDs and prediction scores, which we can add to the query data.  
```{r}
pancrease.query <- pancrease.list[["fluidigmc1"]]
pancrease.anchors <- FindTransferAnchors(reference = pancrease.integrated, query = pancrease.query, dims = 1:30)
predictions <- TransferData(anchorset = pancrease.anchors, refdata = pancrease.integrated$celltype, dims = 1:30)
pancrease.query <- AddMetaData(object = pancrease.query, metadata = predictions)
```

Because we have the original label annotations from our full integrated analysis, we can evaluate how well our predicted cell type annotations match the full reference. In this example, we find that there is a high agreement in cell type classification, which over 97% of cell being labels correctly.  
```{r}
pancrease.query$prediction.match <- pancrease.query$predicted.id == pancrease.query$celltype
table(pancrease.query$prediction.match)
```

```{r}
table(pancrease.query$predicted.id)
```

```{r}
VlnPlot(pancrease.query, c("REG1A", "PPY", "SST", "GHRL", "VWF", "SOX10"), group.by = "predicted.id")
```


```{r}
pancrease.integrated <- merge(pancrease.integrated, pancrease.query)
```


# SCTransform  
Conceptually, this workflow is very similar to what we have previously introduced, where we ‘correct’ (or harmonize) log-normalized expression values across datasets. Here, instead, we will harmonize the Pearson residuals that are output from SCTransform. As demonstrated below, the workflow consists of the following steps:  

- Create a list of Seurat objects to integrate  
- Perform SCTransform normalization separately for each dataset  
- Run the PrepSCTIntegration function on the object list  
- Integrate datasets and proceed with joint analysis  

First, setup the Seurat object list, and run SCTransform on each object separately  
```{r, message=FALSE}
library(Seurat)
library(SeuratData)
library(tidyverse)
library(patchwork)
```

```{r}
options(future.globals.maxSize = 4000 * 1024^2)
data("panc8")

pancrease.list <- SplitObject(object = panc8, split.by = "tech")
pancrease.list <- pancrease.list[c("celseq", "celseq2", "fluidigmc1", "smartseq2")]

for(i in 1:length(pancrease.list)){
    pancrease.list[[i]] <- SCTransform(object = pancrease.list[[i]], verbose = F, show_progress = F)
}
```

Next, select features for downstream integration, and run `PrepSCTIntegration`, which ensures that all necessary Pearson residuals have been calculated.  
```{r}
pancreas.features <- SelectIntegrationFeatures(object.list = pancrease.list, 
                                               nfeatures = 3000)

pancrease.list <- PrepSCTIntegration(object.list = pancrease.list, 
                                     anchor.features = pancreas.features, 
                                     verbose = F)
```

Next, identify anchors and integrate the datasets. Commands are identical to the standard workflow, but make sure to set `normalization.method = 'SCT'`  
```{r}
pancrease.anchors <- FindIntegrationAnchors(object.list = pancrease.list, 
                                            normalization.method = "SCT", 
                                            anchor.features = pancreas.features, 
                                            verbose = F)

pancrease.integrated <- IntegrateData(anchorset = pancrease.anchors, 
                                      normalization.method = "SCT", 
                                      verbose = F)
```

Now proceed with downstream analysis (i.e. visualization, clustering) on the integrated dataset. Commands are identical to the standard workflow, but do not run the `ScaleData` function after integration. You can see that after integration, cells group by their biological cell type (which has been pre-annotated), instead of by their underlying technology.  
```{r}
pancrease.integrated <- RunPCA(object = pancrease.integrated, 
                               verbose = F)

pancrease.integrated <- RunUMAP(object = pancrease.integrated, 
                                dims = 1:30)
```

```{r}
plots <- DimPlot(object = pancrease.integrated, group.by = c("tech", "celltype"))
plots + theme(legend.position = "top") + guides(color = guide_legend(nrow = 3, byrow = TRUE, override.aes = list(size = 3)))
```

























