---
title: "ENCODExplorerData"
author: "yincy"
date: "12/16/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
help(package = "ENCODExplorerData")
```


# ENCODExplorerData: A compilation of metadata from ENCODE  

“The ENCODE (Encyclopedia of DNA Elements) Consortium is an international collaboration of research groups funded by the National Human Genome Research Institute (NHGRI). The goal of ENCODE is to build a comprehensive parts list of functional elements in the human genome, including elements that act at the protein and RNA levels, and regulatory elements that control cells and circumstances in which a gene is active”.  

This package has been designed to facilitate data access by compiling the metadata associated with ENCODE files and making it available in the format of a data table. While this data can be accessed as-is, we recommend using the ENCODExplorer companion package, which contains utility functions for using the online ENCODE search function, downloading selected files, and retrieving control-treatment experimental designs from ENCODE.  

To generate up-to-date version of the data tables exported by this package, simply run the `inst/scripts/make-data.R` script.  



