---
title: "BiocParallel"
author: "yincy"
date: "4/15/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction to BiocParallel  
Numerous approaches are available for parallel computing in R. The CRAN Task View for high performance and parallel computing provides useful high-level summaries and package categorization. http://cran.r-project.org/web/views/HighPerformanceComputing.html Most Task View packages cite or identify one or more of `snow`, `Rmpi`, `multicore` or `foreach` as relevant parallelization infrastructure. Direct support in R for parallel computing started with release 2.14.0 with inclusion of the `parallel` package which contains modified versions of `multicore` and `snow`.  

A basic objective of `BiocParallel` is to reduce the complexity faced when developing and using software that performs parallel computations. With the introduction of the `BiocParallelParam` object, `BiocParallel` aims to provide a unified interface to existing parallel infrastructure where code can be easily executed in different environments. The `BiocParallelParam` specifies the environment of choice as well as computing resources and is invoked by ‘registration‘ or passed as an argument to the `BiocParallel` functions.  

`BioParallel` offers the following conveniences over the 'roll your own' approach to parallel programming.  

- unified interface: `BiocParallelParam` intstances define the method of parallel evaluation (multi-core, snow cluster, etc.) and computing resources (number of workers, error handling, cleanup, etc).  

- parallel iteration over lists, files and vectorized operations: `bplapply`, `bpmapply` and `bpvec` provide parallel list iteration and vectorized operations. `bpiterate` iterates through files distributing chunks to parallel workers.  

- cluster scheduling: When the parallel environment is managed by a cluster scheduler through `batchtools`, job management and result retrieval are considerably simplified.  

- support of `foreach`: The `foreach` and `iterators` packages are fully supported. Registration of the parallel back end uses `BiocParallelParam` instance.  

# Quick start  
```{r}
library(BiocParallel)
```

The test function simply returns the square root of 'x'.  

```{r}
FUN <- function(x){round(sqrt(x), 4)}
```

Functions in `BiocParallel` use the registered back-ends for parallel evaluation. The default is the top entry of the registry list.  

```{r}
registered()
```

Configure your R session to always use a particular back-end configure by setting options named after the back ends in an `.Rprofile` file e.g.,  

```{r}
options(MulticoreParam = quote(MulticoreParam(workers = 4)))
```

When a `BiocParallel` function is invoked with no `BPPARAM` argument the default back-end is used.  

```{r}
system.time(bplapply(1:4, FUN))
```

```{r}
param <- SnowParam(workers = 2, type = "SOCK")
system.time(bplapply(1:4, FUN, BPPARAM = param))
```


# The BiocParallel Interface  
## Classes  
### `BiocParallelParam`  
`BiocParallelParam` instances configure different parallel evaluation environments. Creating or `register()`ing a `Param` allows the same code to be used in different parallel environments without a code-re-write. Params listed are supported on all Unix, Mac and Windows except `MulticoreParam` which is Unix and Mac only.  

- `SerialParam`: Supported on all platforms. Evaluate `BiocParallel`-enabled code with parallel evaluation disabled. This approach is useful when writing new scripts and trying to debug code.  

- `MulticoreParam`: Supported on Unix and Mac. On Windows, `MulticoreParam` dispatches to `SerialParam`. Evaluate `BiocParallel`-enabled code using multiple cores on a single computer. When available, this is the most efficient and least troublesome way to parallelize code. Windows does not support multi-core evaluation (the `MulticoreParam` object can be used, but evaluation is serial). On other operating systems, the default number of workers equals the value of the global option `mc.cores` (e.g., `getOption("mc.cores")`) or, if that is not set, the number of cores returned by `parallel::detectCores() - 2`; when number of cores cannot be determined, the default is 1.  

`MulticoreParam` uses 'forked' processes with 'copy-on-change' semantics - memory is only copied when it is changed. This makes it very efficient to invoke compared to other back-ends. A subtle cost, though, is that R's garbage collector runs periodically, and 'marks' memory as in use. This effectively triggers a copy of the marked memory. R's generational garbage collector is triggered at difficult-to-predict times; the effect in a long-running forked process is that the memory is eventually copied.  

Based on facilities originally implemented in the `multicore` package and subsequently the `parallel` package in base R.  

- `SownParam`: Supported on all platforms. Evaluate `BiocParallel`-enabled code across several distinct R instances, on one or several computers. This is a straightforward approach for executing parallel code on one or serveral computers, and is based on facilities originally implemeted in the `snow` package. Different types of `snow` 'back-ends' are supported, including socket and MPI clusters.  

- `BatchtoolsParam`: Applicable to clusters with formal schedulers. Evaluate `BiocParallel`-enabled code by submitting to a cluster scheduler like SGE.  

- `DoparParam`: Supported on all platforms. Register a parallel back-end supported by the `foreach` package for use with `BiocParallel`.  

The simplest illustration of creating `BiocParallelParam` is  

```{r}
serialParam <- SerialParam()
```

Most parameters have additional arguments influencing behavior, e.g., specifying the number of 'cores' to use when creating a `MulticoreParam` instance.  

```{r}
multicoreParam <- MulticoreParam(workers = 8)
```

### `register()`ing `BiocParallelParam` instances  
The list of registered `BiocParallelParam` instances represents the user’s preferences for different types of back-ends. Individual algorithms may specify a preferred back-end, and different back-ends maybe chosen when parallel evaluation is nested.  

The registry behaves like a ‘stack’ in that the last entry registered is added to the top of the list and becomes the “next used“ (i.e., the default).  

`registered()` invoked with no arguments lists all back-ends.  

```{r}
registered()
```

`bpparam` returns the default form the top of the list.  

```{r}
bpparam()
```

Add a specialized instance with `register`. When `default` is TRUE, the new instance becomes the default.  

```{r}
default <- registered()
register(BatchtoolsParam(workers = 10), default = T)
```

BatchtoolsParam has been moved to the top of the list and is now the default.  

```{r}
registered() %>% names()
```

```{r}
bpparam()
```

Restore the original registry  

```{r}
for (param in rev(default)){
    register(param)
}
```


## Functions
### Parallel looping, vectorized and aggregated operations  
These are used in common functions, implemented as much as possible for all back-ends. The functions include:  

- `bplapply(X, FUN, ...)`: apply in parallel a function `FUN` to each element of `X`. `bplapply` invokes `FUN length(X)` times, each time with a single element of `X`.  

- `bpmapply(FUN, ...)`: Apply in parallel a function `FUN` to the first, second, etc., elements of each argument in ....  

- `bpiterate(ITER, FUN, ...)`: Apply in parallel a function `FUN` to the output of function `ITER`. Data chunks are returned by `ITER` and distributed to parallel workers along with `FUN` and distributed to parallel workers along with `FUN`. Intended for iteration though an undefined number of data chunks.  

- `bpvec(X, FUN, ...)`: Apply in parallel a function `FUN` to subsets of `X`. `bpvec()` invokes function `FUN` as many times as there are cores or cluster nodes, with `FUN` receiving a subset (typically more than 1 element, in contrast to `bplappy`) of `X`.  

- `bpaggregate(x, data, FUN, ...)`: Use the formula in `x` to aggregate `data` using `FUN`.  

### Parallel evaluation environment  






















