---
title: "Inference and analysis of cell-cell communication using CellChat"
author: "YinCY"
format: html
editor: source
---

```{r}
#| message: false
#| warning: false

library(CellChat)
library(patchwork)
library(magrittr)
```

# Introduction
This vignette outlines the steps of inference, analysis and visualization of cell-cell communication network for a single dataset using CellChat. We showcase CellChat’s diverse functionalities by applying it to a scRNA-seq data on cells from lesional (LS, diseased) human skin from patients.

CellChat requires gene expression data of cells as the user input and models the probability of cell-cell communication by integrating gene expression with prior knowledge of the interactions between signaling ligands, receptors and their cofactors.

Upon infering the intercellular communication network, CellChat provides functionality for further data exploration, analysis, and visualization.

# data input, processing and initilization of CellChat object
CellChat requires two user inputs: one is the gene expression data of cells, and the other is the user assigned cell labels.

## Prepare required input data for CellChat analysis
```{r}
load("../../data/CellChat/data_humanSkin_CellChat.rda")
data.input <- data_humanSkin$data
meta <- data_humanSkin$meta

cell_use <- rownames(meta)[meta$condition == "LS"]

data.input <- data.input[, cell_use]
meta <- meta[cell_use, ]
colnames(meta)[1] <- "samples"
```

```{r}
meta$labels %>% unique
```

# Create a CellChat object
```{r}
cellchat <- createCellChat(object = data.input,
                           meta = meta, 
                           group.by = "labels")
cellchat
```


# Set the ligand-receptor interaction database
```{r}
CellChatDB <- CellChatDB.human
CellChatDB.use <- subsetDB(CellChatDB.human)

showDatabaseCategory(CellChatDB.use)
```


# Preprocessing the expression data for cell-cell communication analysis
```{r}
cellchat <- subsetData(cellchat)
cellchat <- identifyOverExpressedGenes(cellchat)
```






















