---
title: "eSet"
author: "yincy"
date: "2/8/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, echo=FALSE}
library(Biobase)
```


# What is an `eSet`?  
1. Coordinate high through-put and phenotype data.  
2. Provide common data container for diverse Bioconductor packages.  

```{r}
getClass("eSet")
```

+ `assayData`: high-throughput data.  
    - efficiently and flexibly contain high-volume data.  
    - each element of list/environment/localEnvironment is a matrix
    - Rows: *features*, e.g., gene names.  
    - Columns: *samples* represented on each chip.  
    - All matricies must have the same dimensions, row names, and column names.  
    - Subclasses determine which matrices *must* be present.  
    
+ `phenoData`: sample covariates.  
    - Contain and document sample covariates.  
    - Structure: AnnotatedDataFrame
    - data: data.frame
        + Rows: sample identifiers
        + Columns: measured covariates
    - varMetadata: data.frame
        + Rows: measured covariate labels
        + Columns: covariate descriptors
        
+ `featureData`: feature covariates.  
    - Purpose: Contain and document feature covariates specific to the experiment, use the `annotation` slot for chip-level descriptions.  
    - Structure: AnnotatedDataFrame
    - data: data.frame
        + Rows: feature identifiers. There match row names of assayData
        + Columns: measured covariates
    - varMetadata: data.frame
        + Rows: measured covariate labels 
        + Columns: covariate descriptors
        
+ `experimentData`: experimental description.  
    - Purpose: Summarize where and how the experiment was performed.  
    - Structure: MIAxE virtual class
        + In Biobase version 2.11 the MIAxE virtual class was introduced as a general container for experiment meta-data. The MIAME class is now subclass of MIAxE. The MIAME class structure is as follows.  
            - title: experiment title
            - name: experiment name(s)
            preprocessing: list of pre-processing steps
            Additional slots
+ `annotation`: assay description.  
    - Purpose: Link experiment to annotation package.
    - Structure: character
        + Label identifying annotation package.  
