---
title: "forestplot"
author: "YinCY"
format: html
editor: source
---

# Introduction
Forest plots date back to 1970s and are most frequently seen in meta-analysis, but are in no way restricted to these. The `forestplot` package is all about providing these in R. It originated form the ‘rmeta’-package’s forestplot function and has a part from generating a standard forest plot, a few interesting features:

- **Text**: 
  + Ability to use a table of text, i.e. the text can consist of several columns if needed. 
  + Ability to use expressions within your text that permit mathematical symbols, e.g. expression(beta)
  + Set the gpar arguments (fontfamily, fontface, cex, etc) for both summary and regular rows. This can be specified down to the each cell.
  
- **Confidence intervals**:
  + Clip confidence intervals to arrows when they exceed specified limits.
  + Multiple confidence bands for the same row.
  + Choose between different estimate markers such as boxes, diamonds, points. 
  + Custom confidence interval drawing functions.
  
- **Legends**：
  + Have a legend on top or to the left of the plot. 
  + Have the legend within the plot’s graph section.
  + Put a box around legend (sharp or rounded corners).
  
- **Others**:
  + Dividing the graph visually by adding horizontal lines.
  + Choose line height to either adapt to viewport (graph) size or specify an exact height in units.
  + Use flexible arguments, you can choose if you want to provide mean, lower, and upper separately or within one array.
  
**Note**: An important difference from the original forestplot is that the current function interprets xlog as the x-axis being in log-format, i.e. you need to provide the data in the antilog/exp format.


# Text
A forest plot is closely connected to text and the ability to customize the text is central.

```{r}
library(forestplot)
library(dplyr)
```

## Table of text
Below is a basic example from the original forestplot function that shows how to use a table of text.

```{r}
base_data <- tibble::tibble(mean  = c(0.578, 0.165, 0.246, 0.700, 0.348, 0.139, 1.017),
                            lower = c(0.372, 0.018, 0.072, 0.333, 0.083, 0.016, 0.365),
                            upper = c(0.898, 1.517, 0.833, 1.474, 1.455, 1.209, 2.831),
                            study = c("Auckland", "Block", "Doran", "Gamsu",
                                      "Morrison", "Papageorgiou", "Tauesch"),
                            deaths_steroid = c("36", "1", "4", "14", "3", "1", "8"),
                            deaths_placebo = c("60", "5", "11", "20", "7", "7", "10"),
                            OR = c("0.58", "0.16", "0.25", "0.70", "0.35", "0.14", "1.02"))

base_data
```

```{r}
base_data %>% 
  forestplot(labeltext = c(study, deaths_steroid, deaths_placebo, OR), clip = c(0.1, 2.5), xlog = T) %>% 
  fp_add_lines() %>% 
  fp_set_style(box = "royalblue", line = "darkblue", summary = "royalblue") %>% 
  fp_add_header(study = c("", "Study"), 
                deaths_steroid = c("Deaths", "(steroid)"), 
                deaths_placebo = c("Deaths", "(placebo)"), 
                OR = c("", "OR")) %>% 
  fp_append_row(mean = 0.531, 
                lower = 0.386, 
                upper = 0.731, 
                study = "Summary",
                OR = "0.53",
                is.summary = TRUE) %>% 
  fp_set_zebra_style(x = "#00F") 

```



























