---
title: "01_Tree_Manipulation"
author: "yincy"
date: "2023-04-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(ape)
library(magrittr)

tr <- read.tree(text = "((a:1,b:1):1,(c:1,d:1):1);")
```


**Tree Manipulation Functions**  
- `drop.tip`: remove one or more tips from a tree.  

- `extract.clade`: keeps all the tips and nodes descending from a node specified by its number or its label.  

- `drop.fossil`: like `drop.tip`, but dropping only the tips that are not contemporaneous with the most distant ones from the root.

- `bind.tree`: build a tree from two trees where the second one is grafted on the first one.

- `rotate`: rotates the internal branch below the most recent common ancestor of a monophyletic group given by the argument group. The resulting tree is equivalent to the original one. This function is convenient when plotting a tree if it is needed to change the order of the tips on the plot.

- `ladderize`: rotates the internal branches of a tree in order to arrange the sister-clades so that the smallest one is on the right-hand side when plotted upwards.  

- `compute.brlen`: modifies or creates the branch lengths of a tree with respect to the second argument, `method`, which may be one of the following.  
    + A character string specifying the method to be used.
    + An R function used to generate random branch lengths.  
    + One or more numeric values(recycled if necessary).  
    
- `compute.brtime`: has a similar action than the previous function but the re- turned tree is ultrametric; the argument method may be:
    + A character string specifying the method to be used to generate the branching times.  
    + A numeric vector of the same length than the number of nodes.

- `is.rooted`  
- `root`  
- `unroot`
- `multi2di`  
- `di2multi`  
- `summary`  
- `is.ultrametric`: all tips euqally distant from the root  
- `balance`: returns, for a fully binary tree, the number of descendants of both sister-lineages from each node.





```{r}
drop.tip(tr, tip = 1:2)
drop.tip(tr, c("a", "b"))
```

```{r}
par(mfrow =c(1,3))
plot(tr)
drop.tip(tr, tip = 1:2, trim.internal = FALSE) %>% plot
drop.tip(tr, tip = 1:2, trim.internal = TRUE) %>% plot
```


It is often convenient to identify tips with numbers, but you must be very carful that many operations are likely to change these numbers.

In all situations, it is safest to identify tips with their labels.

```{r}
t1 <- read.tree(text = "(a:1,b:1):0.5;")
t2 <- read.tree(text = "(c:1,d:1):0.5;")
```

```{r}
par(mfrow = c(2,2))
plot(t1, main = "t1")
plot(t2, main = "t2")
plot(bind.tree(t1, t2), main = "bind t1, t2, position = 0")
plot(bind.tree(t1, t2, position = 0.5), main = "bind t1,t2,position=0.5")
```


```{r}
par(mfrow = c(1, 2))
plot(t1)
plot(rotate(t1, node = 3)) # 3 is the root node
```



# Rooted and Unrooted Trees
```{r}
ta <- read.tree(text = "(a,b,c);")
tb <- read.tree(text = "(a,b,c):;")
tc <- read.tree(text = "((a,b),c);")
td <- read.tree(text = "(a,b,c):0;")
# The presence of a zero `root.edge` allows us to have a rooted tree with a basal trichotomy. 
```

```{r, message=FALSE, warning=FALSE}
par(mfrow = c(2, 2))
plot(ta, main = "ta", type = "cladogram")
plot(tb, main = "tb", type = "cladogram")
plot(tc, main = "tc", type = "cladogram")
plot(td, main = "td", type = "cladogram")
```

```{r}
sapply(list(ta, tb, tc, td), is.rooted)
```


```{r}
td %>% summary
```

```{r}
# all tips equally distant from the root
td %>% is.ultrametric()
```

```{r}
tc %>% balance
```

```{r}
plot(tc)
```


# Case Study
warblers cytochrome b sequences
AJ534526 - AJ534549 and Z73494

```{r}
x <- paste("AJ5345", 26:49, sep = "")
x <- c("Z73494", x)
x
```


```{r}
sylvia.seq <- read.GenBank(access.nb = x)
sylvia.seq
```

```{r}
sylvia.clus <- clustal(sylvia.seq)
```




















