---
title: "ligand-activity-single-cell"
author: "yincy"
date: "4/26/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Single-cell NicheNet's ligand activity analysis  
This vignette shows how NicheNet can be used to predict which ligands might be active in single-cells. If a ligand has a high activity in a cell, this means that target genes of that ligand are stronger expressed in that cell than in other cells. In this example, we will use data from Puram et al. to explore intercellular communication in the tumor microenvironment in head and neck squamous cell carcinoma (HNSCC) (See Puram et al. 2017). More specifically, we will assess the activity of cancer-associated fibroblast (CAF) ligands in malignant cells.  

In order to prioritize ligands regulating a process of interest, you can perform a regression/correlation analysis between ligand activities in cells, and scores of a cell corresponding to the process of interest. For example, in this case study we were interested in finding ligands regulating p-EMT. Therefore we correlated ligand activities to the p-EMT scores of cells.  

The purpose of this single-cell ligand activity analysis is to offer a complementary way to prioritize ligands driving the process of interest and to better analyze heterogeneity in ligand activity between different cells.  

## load packages  
```{r, message=FALSE, warning=FALSE}
library(nichenetr)
library(tidyverse)
```


## Reading in expression data of interacting cells  
First, we will read in the single-cell data from CAF and malignant cells from HNSCC tumors.  
```{r}
hnscc_expression <- readRDS(file = "/home/yincy/git/Data/NicheNet/data/hnscc_expression.rds")
expression <- hnscc_expression$expression
sample_info <- hnscc_expression$sample_info
```

Secondly, we will determine which genes are expressed in CAFs and malignant cells from high quality primary tumors. Therefore, we wil not consider cells from tumor samples of less quality or from lymph node metastases.  
```{r}
tumors_remove <- c("HN10","HN","HN12", "HN13", "HN24", "HN7", "HN8","HN23") 
CAF_ids = sample_info %>% 
  filter(`Lymph node` == 0) %>% 
  filter((tumor %in% tumors_remove == FALSE)) %>% 
  filter(`non-cancer cell type` == "CAF") %>% 
  .$cell
  
malignant_ids = sample_info %>% 
  filter(`Lymph node` == 0) %>% 
  filter(`classified  as cancer cell` == 1) %>% 
  filter((tumor %in% tumors_remove == FALSE)) %>% 
  .$cell


expressed_genes_CAFs = expression[CAF_ids,] %>% 
  apply(2,function(x){10*(2**x - 1)}) %>% 
  apply(2,function(x){log2(mean(x) + 1)}) %>% 
  .[. >= 4] %>% names()

expressed_genes_malignant = expression[malignant_ids,] %>% 
  apply(2,function(x){10*(2**x - 1)}) %>% 
  apply(2,function(x){log2(mean(x) + 1)}) %>% 
  .[. >= 4] %>% names()
```


## load the ligand-target model   
```{r}
ligand_target_matrix <- readRDS(file = "/home/yincy/git/Data/NicheNet/data/ligand_target_matrix.rds")
ligand_target_matrix %>% .[1:5, 1:5]
```

## Perform NicheNet's single-cell liagnd activity analysis  
In a first step, we will define a set of potentially active ligands. As potentially active ligands, we will use ligands that are 1) expressed by CAFs and 2) can bind a (putative) receptor expressed by malignant cells. Putative ligand-receptor links were gathered from NicheNet’s ligand-receptor data sources.  
```{r}
lr_network <- readRDS(file = "/home/yincy/git/Data/NicheNet/data/lr_network.rds")
ligands <- lr_network %>% 
  pull(from) %>% 
  unique()
expressed_ligands <- intersect(expressed_genes_CAFs, ligands)

receptors <- lr_network %>% 
  pull(to) %>% 
  unique()
expressed_receptors <- intersect(expressed_genes_malignant, receptors)

potential_ligands <- lr_network %>% 
  filter(from %in% expressed_ligands, to %in% expressed_receptors) %>% 
  pull(from) %>% 
  unique()
```


```{r}
background_expressed_genes <- expressed_genes_malignant %>% 
  .[. %in% rownames(ligand_target_matrix)]

expression_scaled <- expression %>% 
  .[malignant_ids, background_expressed_genes] %>% 
  scale_quantile()
```

Now perform the ligand activity analysis: infer how well NicheNet’s ligand-target potential scores can predict whether a gene belongs to most strongly expressed genes in a cell compared to other cells. To reduce the running time for this vignette, we will perform the analysis only on 10 example cells from the HN5 tumor. This vignette’s only purpose is to illustrate the analysis.  

In practice, ligand activity analysis for several cells can be better run in parallel (via e.g. parallel::mclapply)!  
```{r}
malignant_hn5_ids <- sample_info %>% 
  filter(tumor == "HN5") %>% 
  filter(`Lymph node` == 0) %>% 
  filter(`classified  as cancer cell` == 1) %>% 
  pull(cell) %>% 
  head(40)

ligand_activities <- predict_single_cell_ligand_activities(
  cell_ids = malignant_hn5_ids, 
  expression_scaled = expression_scaled, 
  ligand_target_matrix = ligand_target_matrix, 
  potential_ligands = potential_ligands
)
```


## Ligand prioritization by regression analysis  
Furthermore, we will also show how you can perform additional analyses by linking the ligand activity in cells to other properties of cells in order to prioritize ligands. As toy example, we will score malignant cells here on the extent to which they express the core p-EMT gene “TGFBI”.  
```{r}
cell_scores_tbl <- tibble(cell = malignant_hn5_ids, 
                          score = expression_scaled[malignant_hn5_ids, "TGFBI"])
```

Then, we will determine the correlation between these p-EMT scores and ligand activities over all cells to prioritize p-EMT-inducing ligands. We hypothesize that ligands might be potential regulators of the p-EMT program if higher ligand activities are associated with higher p-EMT scores. Based on this correlation, we obtained a ranking of potential p-EMT-inducing ligands.  

To do so, we frist need to process and normalize the ligand activities (i.e. pearson correlation values) to make different cells comparable. Here we use modified z-score normalization.  
```{r}
normalized_ligand_activities <- normalize_single_cell_ligand_activities(ligand_activities = ligand_activities)
```

Then, we combine the ligand activities and cell property scores and perform correlation and regression analysis. We can prioritize ligands by ranking them based on the pearson correlation between activity scores and property scores.  
```{r}
output_correlation_analysis <- single_ligand_activity_score_regression(normalized_ligand_activities, cell_scores_tbl)
```

```{r}
output_correlation_analysis %>% 
  arrange(-pearson_regression) %>% 
  select(pearson_regression, ligand)
```

Visualize the relation between ligand activity and the cell's property score of interest  
```{r}
inner_join(cell_scores_tbl, normalized_ligand_activities) %>% 
  ggplot(aes(score, CTGF)) +
  geom_point() +
  geom_smooth(method = "lm")
```


