---
title: "symbol_convertion_from_has_to_mmu"
author: "yincy"
date: "2/25/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Converting NicheNet's model from human to mouse symbols  
One-to-one orthologs were gathered from NCBI HomoloGene and also from ENSEMBL vis biomaRt.  

### Load required packages  
```{r, message=FALSE}
library(nichenetr)
library(tidyverse)
```
 

### Load NicheNet's ligand-receptor model  
```{r}
ligand_target_matrix <- readRDS(file = "/home/yincy/git/Bioconductor/nichenetr/data/ligand_target_matrix.rds")
ligand_target_matrix[1:5, 1:5] # ligands in columns, target genes in rows
```


### Convert the ligand-target model from human to mouse symbols  
Because not all human have a mouse one-to-one ortholog, these genes will be removed from the mouse model.  
```{r}
colnames(ligand_target_matrix) <- ligand_target_matrix %>% 
    colnames() %>% 
    convert_human_to_mouse_symbols()

rownames(ligand_target_matrix) <- ligand_target_matrix %>% 
    rownames() %>% 
    convert_human_to_mouse_symbols()
```


```{r}
ligand_target_matrix <- ligand_target_matrix[!is.na(rownames(ligand_target_matrix)), !is.na(colnames(ligand_target_matrix))]
ligand_target_matrix %>% dim()
```

Show the top 10 targets of TNF (in mouse symbols)  
```{r}
top_targets <- extract_top_n_targets(ligand_oi = "Tnf", top_n = 10, ligand_target_matrix = ligand_target_matrix)
```


convert mouse symbol to human  
```{r}
top_targets %>% names() %>% convert_mouse_to_human_symbols()
```








