---
title: "graph"
author: "yincy"
date: "9/24/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(igraph)
```


# Attributes for Graph Objects
The *graph* pakcage provides representations of graphs (nodes and edges) as S4 classes.  

```{r}
mat <- matrix(c(0, 0, 1, 1, 
                0, 0, 1, 1, 
                1, 1, 0, 1, 
                1, 1, 1, 0), 
              byrow = T, ncol = 4)

rownames(mat) <- letters[1:4]
colnames(mat) <- letters[1:4]
```

```{r}
g1 <- graphAM(adjMat = mat)
plot(g1)
```

## Edge Attributes  
```{r}
edgeDataDefaults(g1)
```

```{r}
edgeDataDefaults(self = g1, attr = "weight") <- 1
edgeDataDefaults(g1, "code") <- "plain"
edgeDataDefaults(g1)
```

```{r}
edgeDataDefaults(g1, "weight")
```


```{r}
edgeData(g1, from = "a", to = "d", attr = "weight")
edgeData(g1, from = "a", attr = "weight")
edgeData(g1, to = "a", attr = "weight")
```

```{r}
allAttrAllEdge <- edgeData(g1)
edgeData(g1, attr = "weight")
```

```{r}
edgeData(g1, from = "a", to = "d", attr = "weight") <- 2
```



# How to use the clusterGraph and distGraph classes  
## clusterGraph  
A *clusterGraph* is a graph defined on a set of nodes that have been clustered or grouped in some fashion. The grouping must form a partition of the nodes. In this graph all nodes within the same cluster are adjacent while there are no edges between clusters.  

Thus, each cluster is a complete graph but there are no between cluster edges.  

```{r}
library(graph)
library(cluster)
data("ruspini")

pm <- pam(ruspini, k = 4)
cG <- new(Class = "clusterGraph", clusters = split(names(pm$clustering), pm$clustering))

nodes(cG)
```

```{r}
library(stats)
km <- kmeans(x = ruspini, centers = 4)
cG.km <- new(Class = "clusterGraph", clusters = split(x = as.character(1:75), 
                                                      f = km$cluster))

inBoth = intersection(cG.km, cG)

inBoth %>% class()
inBoth %>% length()
```

The graph inBoth is of length 1 indicating that there are that many distinct groups. 




















