---
title: "maEndToEnd"
author: "YinCY"
format: html
editor: source
---

# Introduction

In this article we introduce a complete workflow for a typical (Affymetrix) microarray analysis. Data import, preprocessing, differential expression and enrichment analysis are discussed.

The data set used is from a paper studying the differences in gene expression in inflamed and non-inflamed tissue. 14 patients suffering from Ulcerative colitis (UC) and 15 patients with Crohn’s disease (CD) were tested, and from each patient inflamed and non-inflamed colonic mucosa tissue was obtained via a biopsy. This is a typical clinical data set consisting of 58 arrays in total. Our aim is to analyze differential expression (DE) between the tissues. Our results show a substantial overlap with the results of the original paper.

```{r}
#| message: false
#| warning: false

library(Biobase)
library(oligoClasses)
library(ArrayExpress)
library(pd.hugene.1.0.st.v1)
library(hugene10sttranscriptcluster.db)

library(oligo)
library(arrayQualityMetrics)

library(limma)
library(topGO)

library(ggplot2)
library(tidyr)
library(dplyr)
library(janitor)
library(fs)
```

# Import data
```{r}
sdrf <- read.table("/home/yincy/git/data/zhuyi/dermatomyositis/E-MTAB-2967/E-MTAB-2967.sdrf.txt", 
                   sep = "\t", 
                   header = TRUE, 
                   quote = "\"") %>% 
  clean_names()
rownames(sdrf) <- sdrf$array_data_file
sdrf
```

```{r}
files <- list.celfiles("/home/yincy/git/data/zhuyi/dermatomyositis/E-MTAB-2967", 
                       full.names = TRUE)
files %>% str

raw_data <- read.celfiles(filenames = files, 
                          verbose = TRUE)
stopifnot(validObject(raw_data))

es <- ExpressionSet(assayData = exprs(raw_data), 
                    phenoData = AnnotatedDataFrame(sdrf))

validObject(es)
```

```{r}
pData(es) %>% head
```

```{r}
pData(es) <- pData(es)[, c("source_name", "characteristics_individual", "factor_value_disease", "factor_value_phenotype")]
pData(es)
```


# quality control
```{r}
es %>% exprs %>% .[1:5, 1:5]
```

```{r}
exp_raw <- es %>% exprs %>% log2
pcas <- exp_raw %>% t %>% prcomp(scale. = FALSE)

percentVar <- 100 * (pcas$sdev^2 / sum(pcas$sdev^2)) %>% round(digits = 4)

df <- data.frame(PC1 = pcas$x[, 1], 
                 PC2 = pcas$x[, 2], 
                 disease = pData(es)$factor_value_disease, 
                 phenotype = pData(es)$factor_value_phenotype, 
                 individual = pData(es)$characteristics_individual)
df
```

```{r}
df %>% ggplot(aes(PC1, PC2)) +
  geom_point(aes(shape = disease, color = phenotype)) +
  ggtitle("PCA plot of the log-transformed raw expression data") +
  xlab(paste0("PC1, VarExp: ", percentVar[1], "%")) +
  ylab(paste0("PC2, VarExp: ", percentVar[2], "%")) +
  scale_shape_manual(values = c(4, 15)) +
  scale_color_manual(values = c("darkorange2", "dodgerblue4"))
```

```{r}
oligo::boxplot(raw_data, 
               target = "core", 
               main = "Boxplot of log2-intensites for the raw data")
```

```{r}
arrayQualityMetrics::arrayQualityMetrics(es, 
                                         outdir = "./", 
                                         force = TRUE, 
                                         do.logtransform = TRUE, 
                                         intgroup = c("factor_value_disease"))
```





























