---
title: "S4"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load package, message=FALSE}
library(ALL)
library(GenomicRanges)
library(Biobase)
```


## Overview  
The S4 system in R is a system for object oriented programing. Confusingly, R has support for at least 3 different systems for object oriented programming: S3, S4 and S5 (also known as reference classes).  


The S4 system is heavily used in Bioconductor, whereas it is very lightly used in “traditional” R and in packages from CRAN. As a user it can be useful to recognize S4 objects and to learn some facts about how to explore, manipulate and use the help system when encountering S4 classes and methods.  


## S3 and S4 classes  
S3  
```{r}
df <- data.frame(y = rnorm(10), x = rnorm(10))
lm.object <- lm(y ~ x, data = df)
```

```{r}
names(lm.object)
```

```{r}
class(lm.object)
```

In standard R, an S3 object is essentially a list with a class attribute on it. The problem with S3 is that we can assign any class to any list, which is nonsense.   

```{r}
xx <- list(a = letters[1:3], b = rnorm(3))
xx
```
```{r}
class(xx) <- "lm"
xx
```

S4  
```{r}
data(ALL)
```

```{r}
class(ALL)
isS4(ALL)
```


## Constructors and getting help  
The Bioconductor coding standards suggests that an S4 class should have a name that begins with a capital letter and a constructor function with the same name as the class.  

It is common that the constructor function is documented on the same help page as the class.  

An example of a class in Bioconductor that does not have a constructor function is the BSParams class from BSgenome used for constructing calls to the bsapply function (applying functions over whole genomes).  


## Slots and accessor functions  
you can get the class definition as  
```{r}
getClass("ExpressionSet")
```

In this out put you'll see two things  
- A number of `slots` are mentioned together with a name and a `class`.  
- The class "extends" the class `eSet` directly.  


Data inside an S4 class are organized into slots. You access slots by using either ‘@’ or the ’slots()` function  
```{r}
ALL@annotation
```

```{r}
slot(ALL, "annotation")
```
 
However, as a user you **should never have to access slots directly**. This is important to understand. You should get data out of the class using “accessor” functions. Frequently accessor functions are named as the slot or perhaps get and the slot name.  

```{r}
annotation(ALL)
```

(the `get` version of this name is `getAnnotation()` - different package authors use different styles). Not all slots have an accessor function, because slots may contain data which is not useful to the user.

Traditionally, accessor functions are documented on the same help page as the class itself.  

Accessor functions does not always precisely refer to a slot. For example, for ExpressionSet we use exprs() to get the expression matrix, but there is no slot called exprs in the class definition. We still refer to exprs() as an accessor function.  

By only using accessor functions you are protecting yourself (and your code) against future changes in the class definition; accessor functions should always work.  


## Class inheritance  
Class inheritance is used a lot in Bioconductor. Class inheritance is used when you define a new class which “is almost like this other class but with a little twist”. For example ExpressionSet inherits from eSet, and when you look at the class definition you cannot easily see a difference. The difference is that ExpressionSet is meant to contain expression data and has the exprs() accessor.  


## Outdated S4 classes  
```
new_object <- updateObject(old_object)
```


## S4 Methods  
You can think of S4 methods as simple functions. A method is a function whichh can look at its arguments and decide what to do. One way to mimic a method is by a function definition like the following  
```
mimicMethod <- function(x){
  if(is(x, "matrix"))
    methods1(x)
  if(is(x, "data.frme"))
    methods2(x)
  if(is(x, "IRanges"))
    methods3(x)
}
```
This function examines the x argument and runs different sets of code (method1, method2, method3) depending on which class x is.  
```{r}
showMethods("as.data.frame")
```

The different values of `x` here are called 'signatures'.  

Actually, this does not show you the actual methods, it just shows you which values of x a method has been defined for. To see the code  
```{r}
getMethod(f = "as.data.frame", "DataFrame")
```

Lingo - `as.data.frame` is a generic method. It operatures on different signatures (values of x) and each signature has an associated method. This method is said to “dispatch” on `x`.   


```{r}
showMethods("findOverlaps")
getMethod("findOverlaps", signature(query = "GenomicRanges", subject = "GenomicRanges"))
```

  










