---
title: "RSQLite"
author: "yincy"
date: "3/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(DBI)
library(RSQLite)
```

## Creating a new database  
to create a new SQLite database, you simply supply the filename to `dbConnect()`  
```{r}
mydb <- dbConnect(drv = RSQLite::SQLite(), "my-db.sqlite")
dbDisconnect(mydb)
unlink(x = "my-db.sqlite")
```

if you just need a temporary database, use either `""` (for an on-disk database) or `":memory:"` or `"file:memory:"`(for a in-memory database). This database will be automatically delected when you disconnect from it.  
```{r}
mydb <- dbConnect(drv = RSQLite::SQLite(), "")
dbDisconnect(mydb)
```

## Loading data  
You can easily copy an R data frame into a SQLite database with `dbWriteTable()`  
```{r}
mydb <- dbConnect(drv = RSQLite::SQLite(), "")
dbWriteTable(conn = mydb, name = "mtcars", value = mtcars)
dbWriteTable(conn = mydb, name = "iris", value = iris)
dbListTables(mydb)
```

## Queries  
Issue a query with `dbGetQuery()`  
```{r}
qr <- dbGetQuery(conn = mydb, statement = "select * from mtcars limit 5")
```

Not all R variable names are valid SQL variable names, so you may need to escape them with `"`  
```{r}
qr <- dbGetQuery(conn = mydb, statement = 'select * from iris where "Sepal.Length" < 4.6')
```

If you need to insert the value from a user into a query, don’t use paste()! That makes it easy for a malicious attacker to insert SQL that might damager your database or reveal sensitive information. Instead, use a parameterised query:  
```{r}
qr <- dbGetQuery(conn = mydb, statement = 'select * from iris where "Sepal.Length" < :x', 
           params = list(x = 4.6))
```

## Batched queries  
If you run a query and the results don’t fit in memory, you can use dbSendQuery(), dbFetch() and dbClearResults() to retrieve the results in batches. By default dbFetch() will retrieve all available rows: use n to set the maximum number of rows to return.  
```{r}
rs <- dbSendQuery(conn = mydb, statement = "select * from mtcars")
while(!dbHasCompleted(rs)){
    df <- dbFetch(res = rs, n = 10)
    print(nrow(df))
}
dbClearResult(rs)
```


## Multiple parameterised queries  
You can use the same approach to run the same parameterised query with different parameters. Call dbBind() to set the parameters:  
```{r}
rs <- dbSendQuery(conn = mydb, 'select * from iris where "Sepal.Length" < :x')
dbBind(res = rs, params = list(x = 4.5))
nrow(dbFetch(rs))

dbBind(rs, params = list(x = 4))
nrow(dbFetch(rs))
dbClearResult(rs)
```

you can pass multiple parameters in one call to `dbBind()`  
```{r}
rs <- dbSendQuery(conn = mydb, statement = 'select * from iris where "Sepal.Length" = :x')
dbBind(res = rs, params = list(x = seq(4, 4.4, by = 0.1)))
nrow(dbFetch(rs))
dbClearResult(rs)
```


## Statements  
DBI has new functions dbSendStatement() and dbExecute(), which are the counterparts of dbSendQuery() and dbGetQuery() for SQL statements that do not return a tabular result, such as inserting records into a table, updating a table, or setting engine parameters. It is good practice, although currently not enforced, to use the new functions when you don’t expect a result.  
```{r}
dbExecute(mydb, statement = 'delete from iris where "Sepal.Length" < :x')
rs <- dbSendStatement(mydb, 'delete from iris where "Sepal.Length" < :x')
dbBind(rs, params = list(x = 4.5))
dbGetRowsAffected(rs)
dbClearResult(rs)
```

