---
title: "ImmG_RNA-seq_cell_info"
author: "yincy"
date: "11/28/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
```

```{r}
library(GEOquery)
library(magrittr)
```


These RNAseq profiles were generated by ImmGen labs in a combined study associating RNAseq and ATACseq performed on cell populations sorted in parallel (companion ATACseq datasets are found in GSE100738). The 98 cell populations include all adaptive and innate lymphocytes (B, abT, gdT, Innate-Like Lymphocytes), myeloid cells (dendritic cells, macrophages, monocytes), mast cells and neutrophils. Most were prepared from baseline unchallenged mice, some after cell activation (LPS, anti-CD3, viral infection). For B and T lymphocytes, many successive steps of their known differentiation cascades in the thymus and bone marrow are included.  

```{r, warning=FALSE}
gse109125 <- getGEO(GEO = "GSE109125", GSEMatrix = F)
```


```{r}
gse109125_header <- slot(gse109125, "header")
gse109125_gsms <- slot(gse109125, "gsms")
```

```{r}
gse109125_gsms %>% names()
```

```{r}
gse109125_gsms %>% class()
```


```{r}

```

```{r}
cells <- sapply(gse109125_gsms, function(x){
    cell <- c(x@header$characteristics_ch1[1],
              x@header$characteristics_ch1[2],
              x@header$characteristics_ch1[3],
              x@header$characteristics_ch1[4],
              x@header$characteristics_ch1[5],
              x@header$description[1],
              x@header$geo_accession,
              x@header$instrument_model,
              x@header$source_name_ch1,
              x@header$title)
}, simplify = T)
```



```{r}
gse109125_gsms$GSM2932488@header
```



























