---
title: "S4Vectors"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
```

```{r}
library(S4Vectors)
```


### Vector-like and list-like objects  
In the context of the `S4Vectors` package, a vector-like object is an ordered finite collection of elements.   

All vector-like objects have three main properties:   
    1. a notion of length or number of elements   
    2. the ability to extract elements to create new vector-like objects  
    3. the ability to be concatenated with one or more vector-like objects to form larger vectorlike objects.   
    
The main functions for these three operations are `length`, `[`, and `c`.   

Some vector-like objects can also have a list-like semantic, which means that individual
elements can be extracted with `[[`.  

In `S4Vectors` and many other Bioconductor packages, vector-like and list-like objects derive from the `Vector` and `List` virtual classes, respectively. **Note that List is a subclass of Vector**.  


#### Vector-like objects  
The Rle class defined in the `S4Vectors` package is used to represent a run-length encoded (compressed) sequence of *logical*, *integer*, *numeric*, *complex*, *character*, *raw*, or *factor* values.  
```{r}
showClass("Rle")
```

```{r}
set.seed(0)
lambda <- c(rep(0.001, 4500), seq(0.001, 10, length = 500), seq(10, 0.001, length = 500))

xVector <- rpois(n = 1e7, lambda = lambda)
yVector <- rpois(n = 1e7, lambda = lambda[c(251:length(lambda), 1:250)])

xRle <- Rle(values = xVector)
yRle <- Rle(values = yVector)
```


```{r}
length(xRle)
xRle[1]
```

```{r}
zRle <- c(xRle, yRle)
```


#### Subsetting a vector-like object  
When this subsetting does not duplicate or reorder the elements being extracted, the result is
called a *subsequence*.  

the S4Vectors package supports seven additional functions for sequence extraction:  
    1. `window` - Extracts a subsequence over a specified region.   
    2. `subset` - Extracts the subsequence specified by a logical vector.  
    3. `head` - Extracts a consecutive subsequence containing the first n elements.  
    4. `tail` - Extracts a consecutive subsequence containing the last n elements.  
    5. `rev` - Creates a new sequence with the elements in the reverse order.  
    6. `rep` - Creates a new sequence by repeating sequence elements.  
    
```{r}
xSnippet <- window(xRle, 4751, 4760)
```

```{r}
head(xSnippet)
```

```{r}
tail(xSnippet)
```

```{r}
rev(xSnippet)
```


```{r}
rep(xSnippet, 2)
```

```{r}
subset(xSnippet, subset = xSnippet >= 5   )
```

#### Concatenating vector-like objects  
The `S4Vectors` package uses two generic functions, `c` and `append`, for concatenating two Vector derivates.  
```{r}
c(xSnippet, rev(xSnippet))
```


```{r}
append(xSnippet, xSnippet, after = 3)
```


#### Looping over subsequences of vector-like objects  
In R, for looping can be an expensive operation. To compensate for this, the S4Vectors package provides `aggregate` and `shiftApply` methods (`shiftApply` is a new generic function defined in `S4Vectors`) to perform calculations over subsequences of vector-like objects.  

The `aggregate` function combines sequence extraction functionality of the window function with looping capabilities of the `sapply` function.  

```{r}
xSnippet
```


```{r}
aggregate(xSnippet, start = 1:8, width = 3, FUN = median)
```

The shiftApply function is a looping operation involving two vector-like objects whose elements are lined up via a positional shift operation.   

If we did not know the size of the
shift, we could estimate it by finding the shift that maximizes the correlation between xRle and yRle.  

```{r}
cor(xRle, yRle)
```

```{r}
shifts <- seq(from = 235, to = 265, by = 1)
```

```{r}
corrs <- shiftApply(SHIFT = shifts, yRle, xRle, FUN = cor)
```

```{r}
plot(shifts, corrs)
```


The functions `runValue` and `runLength` extract the run values and run lengths from an Rle object respectively.  

```{r}
head(runValue(xRle))
head(runLength(xRle))
```

The Rle class supports many of the basic methods associated with R atomic vectors including the `Ops`, `Math`, `Math2`, `Summary`, and `Complex` group generics.  

```{r}
xRle > 0
```

```{r}
xRle + yRle
```

```{r}
xRle > 0 | yRle > 0 
```


```{r}
range(xRle)
```

```{r}
summary(xRle > 0 | yRle > 0 )
```


```{r}
log1p(xRle)
```

```{r}
cor(xRle, yRle)
```

```{r}
shiftApply(249:251, yRle, xRle, FUN = function(x, y){var(x, y) / (sd(x)*sd(y))})
```


### List-like objects  
Just as with ordinary R list objects, List-derived objects support `[[` for element extraction, `c` for concatenating, and `lapply`/`sapply` for looping.  


the `S4Vectors` package introduces the `endoapply` function to perform an endomorphism equivalent to `lapply`, i.e. it returns a List derivative of the same class as the input rather than a list object.  


```{r}
showClass("DataFrame")
```

```{r}
df <- DataFrame(x = xRle, y = yRle)
```

```{r}
sapply(df, class)
```


```{r}
sapply(df, summary)
```


```{r}
sapply(as.data.frame(df), summary)
```

```{r}
endoapply(X = df, FUN = "+", 0.5)
```


### Vector Annotations  
Often when one has a collection of objects, there is a need to attach metadata that describes the collection in some way. Two kinds of metadata can be attached to a Vector object.  
    1. Metadata about the object as a whole: this metadata us accessed via `metadata` accessor and is represented as an ordinary **list**.  
    
    2. Metadata about the individual elements of the object: this metadata is accessed via the `mcols` accessor (`mcols` stands for *metadata columns*) and is represneted as a DataTable object. This DataTable object can be thought of as hte rsult of binding together one or several vector-like objects (the metadata columns) of the same length as the Vector object. Each row of the DataTable object annotates the coresponding element of the Vector object.  
   
   
 

`aggregate`: Splits the data into subsets, computes summary statistics for each, and returns the result in a convenient form.  
```{r}
rl <- Rle(c(1,1,1,1,2,2,3,3,2,2))
ir <- IRanges(start = c(2, 6), width = 2)
aggregate(x = rl, by = ir, FUN = mean)
```


```{r}
coverage(ir)
```

```{r}
ir <- IRanges(start = 1:10, width = 3)
rl <- coverage(ir)
```

select regions by `slice()` function  
```{r}
slice(x = rl, lower = 2, upper = 3, includeLower = T, includeUpper = T)
```






