---
title: "Inferring ligands-to-target signaling paths"
author: "yincy"
date: "3/3/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Infer signaling paths between ligand(s) and target(s) of interest  
To determine signaling paths between a ligand and target of interest, we look at which transcription factors are best regulating the target genes and are most closely downstream of the ligand (based on the weights of the edges in the integrated ligand-signaling and gene regulatory networks).  

Then, the shortest paths between these transcription factors and the ligand of interests are determined and genes forming part in this path are considered as important signaling mediators.  

Finally, we look in our collected data source networks for all interactions between the ligand, signaling mediators, transcription factors and target genes. This allows to both prioritize signaling mediators and check which of all collected data sources support the ligand-target predictions of interest.  

For this analysis, you need to define:  

- one or more ligands of interest  
- one or more target genes of interest  

In this vignette, we will demonstrate how to infer signaling paths between a CAF-ligand (CAF = cancer-associated fibroblast) of interest and some of its top-predicted p-EMT target genes.  

```{r load required packages, message=FALSE}
library(nichenetr)
library(tidyverse)
library(igraph)
```


```{r load required data, message=FALSE}
weighted_network <- readRDS(file = "/home/yincy/git/Data/NicheNet/data/weighted_networks.rds")
ligand_tf_matrix <- readRDS(file = "/home/yincy/git/Data/NicheNet/data/ligand_tf_matrix.rds")

lr_network <- readRDS(file = "/home/yincy/git/Data/NicheNet/data/lr_network.rds")
sig_network <- readRDS(file = "/home/yincy/git/Data/NicheNet/data/signaling_network.rds")
gr_network <- readRDS(file = "/home/yincy/git/Data/NicheNet/data/gr_network.rds")
```


as example, we will infer signaling paths between the CAF-ligand TGFB3 and its top-predicted p-EMT target genes TGFBl, LAMC2 and TNC.  
```{r}
ligands_all <- "TGFB3"
# this can be a list of multiple ligands if reuired  

targets_all <- c("TGFBI", "LAMC2", "TNC")

active_signaling_network <- get_ligand_signaling_path(
    ligand_tf_matrix = ligand_tf_matrix, 
    ligands_all = ligands_all, 
    targets_all = targets_all, 
    weighted_networks = weighted_network
)
```


For better visualization of edge weights: normalize edge weights to make them comparable between signaling and gene regulatory interactions  
```{r}
active_signaling_network_min_max <- active_signaling_network

active_signaling_network_min_max$sig <- active_signaling_network_min_max$sig %>% 
    mutate(weight = (weight - min(weight)) / (max(weight) - min(weight)) + 0.75)

active_signaling_network_min_max$gr <- active_signaling_network_min_max$gr %>% 
    mutate(weight = (weight - min(weight)) / (max(weight) - min(weight)) + 0.75)
```

```{r}
graph_min_max <- diagrammer_format_signaling_graph(
    signaling_graph_list = active_signaling_network_min_max, 
    ligands_all = ligands_all, 
    targets_all = targets_all, 
    gr_color = "steelblue", 
    sig_color = "indianred"
)

DiagrammeR::render_graph(graph_min_max, layout = "tree")
```


```{r visualization with igraph, message=FALSE}
path_to_vis <- bind_rows(active_signaling_network_min_max$sig, active_signaling_network_min_max$gr)

path_to_vis <- aggregate(path_to_vis[, 3], by = path_to_vis[, 1:2], sum)
net <- graph_from_data_frame(d = path_to_vis)

vcol <- grep("grey40", vcount(net))
vcol[V(net)$name == "TGFB3"] <- "tomato"
vcol[V(net)$name %in% c("TGFBI", "LAMC2", "TNC")] <- "steelblue"
vcol[is.na(vcol)] <- "grey70"


l <- layout_with_fr(net)
l <- norm_coords(l)
```

```{r}
par(mar = c(0, 0, 0, 0), bg = "black")
plot(net, 
     vertex.color = vcol, 
     layout = l * 1.8, 
     edge.width = E(net)$weight * 2, 
     edge.arrow.size = 0.8, 
     vertex.label.dist = 0, 
     vertex.size = 30, 
     vertex.label.cex = 0.5, 
     vertex.shape = "circle", 
     rescale = F, 
     xlim = c(-2, 2), 
     ylim = c(-2, 2), 
     vertex.label.color = "white")
```

 
 **Look at which collected data sources support the interactions in this network**  
```{r}
data_source_network <- infer_supporting_datasources(
    signaling_graph_list = active_signaling_network_min_max, 
    lr_network = lr_network, 
    sig_network = sig_network, 
    gr_network = gr_network
)
```











